---
title: "D2: Reconciling the biomass measurements, and testing both $N_0$ back-calculations"
subtitle: "Does the published Pseudomonas CUE result change under a better-supported $N_0$?"
author: "OxygenModel D-series"
date: today
date-format: "YYYY-MM-DD"
format:
  pdf:
    documentclass: scrartcl
    papersize: a4
    geometry:
      - top=22mm
      - bottom=22mm
      - left=20mm
      - right=20mm
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    fig-pos: "H"
    include-in-header: ../_shared/header.tex
bibliography: ../_shared/references.bib
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "pdf"
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse); library(knitr); library(jsonlite)
})

BASE <- normalizePath("../..", mustWork = TRUE)
CMP  <- file.path(BASE, "runs", "D2_analysis")
rd   <- function(f) suppressWarnings(suppressMessages(
  readr::read_csv(file.path(CMP, f), show_col_types = FALSE, progress = FALSE)))
VERS <- jsonlite::fromJSON(file.path(BASE, "env", "versions.json"))

kb <- function(x, ...) knitr::kable(x, booktabs = TRUE, linesep = "", ...)
kbs <- function(x, caption = NULL, ...) {
  if (!is.null(caption)) {
    caption <- gsub("([#$%&_{}])", "\\\\\\1", caption)
    caption <- gsub("\\^", "\\\\textasciicircum{}", caption)
    caption <- gsub("~", "\\\\textasciitilde{}", caption)
  }
  k <- knitr::kable(x, booktabs = TRUE, linesep = "", format = "latex",
                    caption = caption, ...)
  knitr::asis_output(paste0("\n\\begingroup\\footnotesize\n",
                            paste(k, collapse = "\n"), "\n\\endgroup\n"))
}
sci <- function(x, d = 2) formatC(x, format = "e", digits = d)

a1m  <- rd("A1_CellCounts_is_taxon_mean.csv")
a1c  <- rd("A1_CellCounts_vs_ODFC_columns.csv")
a2r  <- rd("A2_three_way_density_per_replicate.csv")
a2t  <- rd("A2_three_way_density_by_taxon.csv")
a2s  <- rd("A2_pairwise_ratio_summary.csv")
a2a  <- rd("A2_conversion_assumptions.csv")
a3p  <- rd("A3_Ninoc_provenance_candidates.csv")
a3s  <- rd("A3_Ninoc_structure_by_taxon.csv")
a4   <- rd("A4_agreement_published_vs_perrep.csv")
b1   <- rd("B1_exponential_growth_check.csv")
b1s  <- rd("B1_growth_check_summary.csv") |> tibble::deframe()
b1t  <- rd("B1_growth_check_by_taxon.csv")
b2   <- rd("B2_N0_both_routes_per_replicate.csv")
b2s  <- rd("B2_N0_ratio_summary.csv") |> tibble::deframe()
b2t  <- rd("B2_N0_ratio_by_taxon.csv")
b2r  <- rd("B2_N0_ratio_vs_r.csv")
b3t  <- rd("B3_R_and_CUE_by_taxon.csv")
b3e  <- rd("B3_effect_sizes_vs_window_choice.csv")
c1   <- rd("C1_thermal_parameters_three_routes.csv")
c2   <- rd("C2_per_temperature_three_routes.csv")
c3   <- rd("C3_growth_carbon_Topt_three_routes.csv")
c4   <- rd("C4_optima_three_routes.csv")
c4v  <- rd("C4_conclusion_holds_three_routes.csv")
PSEUDO_SCALE <- as.numeric(readLines(file.path(CMP, "C0_pseudomonas_scale.txt")))

# One accessor over both summary vectors; keys are unique between them.
g <- function(v) as.numeric(if (v %in% names(b2s)) b2s[[v]] else b1s[[v]])
rat <- function(k) a2s$median[a2s$ratio == k]
```

# Subsequent developments {.unnumbered}

::: {.callout-important title="Added 2026-08-07 — read before quoting any absolute number below"}
This report was written against commit `ed5e1a6`. It is the record of what was
found **at that time**; it has not been re-run. Three things changed on `main`
afterwards, and this preamble was added without re-rendering the analysis.

1. **The backward route is now the default.** `dc0951c` (2026-08-06) introduced
   depletion-anchored $N_0$ and `config.R` ships `N0_METHOD <- "depletion"` —
   the route this report argues for. Its recommendation is implemented, not
   pending. The forward route survives as the `"initial"` fallback.

2. **`FC_TO_CELLS_PER_L` was re-derived.** `1cb16fb` (2026-08-07) replaced
   `909916` — a placeholder tuned to a target median $R$, implying an impossible
   sub-1× dilution — with `10255100`, derived from the stated dilution chain.
   This sets the **absolute** respiration and CUE scale only. **The absolute
   per-cell respiration figures in this report predate that change and should
   not be quoted as current.** Relative results and every conclusion about
   *ordering* or *optima* are unaffected.

3. **`data/Ninoc.csv` was rewritten, but its values did not move.** `5ed17b0`
   (2026-08-03) rewrites all 75 rows; the median `N_inoculation_cells_per_L`
   shifts only from 24,605,191 to 24,788,014 cells/L (median row-wise ratio
   1.000×). **The central finding below — that the two independent density
   measurements agree with each other and both sit ~49–66× above `N_inoc` —
   is therefore not resolved by that rewrite and stands as current.**

Recomputing under the new constant and default route is out of scope for the
documentation change that added this note, and has not been done.
:::

# Verdict {.unnumbered}

```{=latex}
\begin{verdictbox}
```

**The published temperature result does not change. The absolute respiration
and CUE values do — by a factor of about 57.**

Two independent measurements of the cells in a vial — the recorded $\mathrm{OD}_{600}$ and
the flow-cytometry count — agree with each other to a median factor of
`r sprintf("%.2f", rat("ratio_FC_over_measuredOD"))`. The quantity the pipeline
actually uses, `N_inoculation_cells_per_L`, agrees with neither: it is a median
**`r sprintf("%.0f", rat("ratio_FC_over_Ninoc"))`×** below the flow-cytometry
count and **`r sprintf("%.0f", rat("ratio_measuredOD_over_Ninoc"))`×** below the
OD-derived count. Because $R = K\,O_{2,\text{ref}} / N_0$, per-cell respiration
is high by that factor.

The shared assumption of both back-calculations survives testing:
flow-cytometry fold change matches $\exp(r\,\Delta t)$ at the oxygen-derived $r$
to a median of **`r sprintf("%.3f", g("ratio median"))`**, with
`r as.integer(g("n within a factor of 2"))` of `r as.integer(g("n replicates"))`
replicates inside a factor of two. So the forward route's $\exp(r\delta)$
correction is the right *size* (median `r sprintf("%.2f", g("forward amplification exp(r*delta) median"))`×);
it is the starting value that is wrong. Backward over forward has median
**`r sprintf("%.0f", g("ratio backward/forward MEDIAN"))`**
(range `r sprintf("%.0f", g("ratio backward/forward min"))`–`r sprintf("%.0f", g("ratio backward/forward max"))`),
and all `r as.integer(g("n kept for the backward route"))` retained replicates
exceed 10. For scale, D1 measured the fit-window choice as worth at most a
factor of 1.23 on $R$.

**But the published *Pseudomonas* temperature conclusion is untouched.** Growth
$T_{opt}$ (`r sprintf("%.2f", c4$Topt_growth_rate_C[c4$route=="forward"])` °C)
and respiration $T_{opt}$ (`r sprintf("%.2f", c4$Topt_respiration_C[c4$route=="forward"])` °C)
are identical to eight significant figures under all three routes, because
neither quantity contains $N_0$. The CUE optimum moves from
**`r sprintf("%.2f", c4$Topt_CUE_C[c4$route=="forward"])` °C** to
**`r sprintf("%.2f", c4$Topt_CUE_C[c4$route=="backward"])` °C** — a shift of
`r sprintf("%.2f", abs(c4$shift_vs_D1_CUE[c4$route=="backward"]))` °C — and
stays 1.5–1.7 °C below the growth optimum, inside the stated 30–35 °C band. The
paper's stated conclusion holds under every route tested (§4.4).

**One correction to the brief.** `06_main_figures.R` does *not* use the
taxon-mean initial count. It reads `data/OD_r_FC_r.csv`, whose `FC_Initial` is
per-replicate; the recomputation reproduces D1's `r_FC` to exactly 0. Using the
taxon mean instead would *degrade* the published validation ($R^2$
`r sprintf("%.3f", a4$R2_lmer_fitted[a4$variant=="r_O2 ~ r_FC, per-replicate initial (AS PUBLISHED)"])`
→ `r sprintf("%.3f", a4$R2_lmer_fitted[a4$variant=="r_O2 ~ r_FC, taxon-mean initial (COUNTERFACTUAL)"])`).
The published choice is the better one.

```{=latex}
\end{verdictbox}
```

# What this report does, and what it does not

Per-cell respiration is $R = K\,O_{2,\text{ref}} / N_0$, where $N_0$ is the cell
density at the start of the fitted oxygen trajectory. In a sealed, headspace-free
vial that density cannot be measured without destroying the closed-system
measurement, so it is back-calculated. This is a considered design choice, not a
defect; the question is only which back-calculation the data support.

`config.R` sets `N0_BACKPROJECT <- TRUE`, giving

$$N_0^{\text{forward}} = N_{\text{inoc}} \cdot e^{r\delta}
\qquad\Longrightarrow\qquad
\log R = \log K + \log O_{2,\text{ref}} - \log N_{\text{inoc}} - r\delta$$

so respiration is an explicit *decreasing* function of the fitted growth rate.
The alternative uses the endpoint count from the same vial:

$$N_0^{\text{backward}} = N_{\text{end}} \cdot e^{-r\,(t_{\text{end}} - t_0)}$$

Both assume exponential growth — the forward route across the roughly 45 min
stabilisation period the paper itself says the oxygen signal cannot be trusted
in, the backward route only across the interval the oxygen model already assumes
it over. In the *Candida* application of this method the equivalent term could
only be **bounded**, because no endpoint biomass existed. Here it can be
**measured**.

**Nothing was changed.** `N0_BACKPROJECT` remains `TRUE`, the QC thresholds, fit
windows, exclusions and cell sizes are untouched. `10_temperature_cue.R` gained
an `OXYMODEL_N0_ROUTE` switch on the same pattern as D1's `OXYMODEL_RESULTS`,
defaulting to the published behaviour; the forward run reproduces
`runs/D1_baseline` byte-identically. `results/`, `data/` and `runs/D1_baseline/`
were verified unchanged by checksum (57 of 57 sums OK, 0 failed).

# Part A — the three statements of cell density

## The conversions, and every assumption in them

```{r a2-assum}
a2a |> dplyr::transmute(Quantity = quantity, Value = value,
                        Source = stringr::str_trunc(source, 62)) |>
  kbs(caption = "Every conversion assumption. The two that are ASSUMED rather than stated are flagged as such.")
```

The flow-cytometry dilution chain, shown as arithmetic:

$$\underbrace{\frac{500}{490}}_{\text{glutaraldehyde}} \times
  \underbrace{\frac{500}{50}}_{\text{PBS}} \times
  \underbrace{\frac{502.5}{500}}_{\text{SYBR}} = 10.2551$$

so an original culture density is the CytoFLEX reading × 10.2551 × 10⁶ cells L$^{-1}$,
**on the assumption that the instrument reports events per $\mu$L of the measured,
diluted sample**. The paper does not state the output unit; this is the
instrument default and is the largest single assumption in Part A.

No taxon-specific $\mathrm{OD}_{600}$-to-cell factor is published for these 15 environmental
isolates. The generic E. coli reference, 8 × 10⁸ cells mL$^{-1}$ per OD unit
(documented spread 5 × 10⁸ – 2 × 10⁹), is used for all of them, and a range is
carried rather than a point estimate. **The OD route needs no dilution-chain
assumption at all**, which is what makes the agreement below load-bearing.

## The three-way table

```{r a2-taxon}
a2t |>
  dplyr::transmute(Taxon,
                   `Measured OD` = sci(measuredOD_cells_per_L_med, 2),
                   `Flow cyt.` = sci(FC_cells_per_L_med, 2),
                   `N_inoc` = sci(N_inoc_med, 2),
                   `FC / N_inoc` = sprintf("%.0f", r_FCoverNinoc),
                   `OD / N_inoc` = sprintf("%.0f", r_measODoverNinoc),
                   `FC / OD` = sprintf("%.2f", r_FCovermeasOD)) |>
  kbs(caption = "Cell density at inoculation, cells per litre, taxon medians. The stated-OD column (4.00e+08 for every taxon by construction) and the median OD600 are omitted for width. The last three columns are the pairwise ratios.",
      align = "lrrrrrr")
```

```{r a2-ratios}
a2s |>
  dplyr::transmute(Ratio = stringr::str_remove(ratio, "^ratio_"),
                   n, Min = sprintf("%.2f", min), `25th` = sprintf("%.2f", q25),
                   Median = sprintf("%.2f", median), `75th` = sprintf("%.2f", q75),
                   Max = sprintf("%.2f", max),
                   `Fold spread` = sprintf("%.0f", fold_spread)) |>
  kbs(caption = "Pairwise ratios across all 75 replicates.", align = "lrrrrrrr")
```

![The three statements of cell density, per taxon and replicate. The two measurements overlap; the value the pipeline uses sits roughly two decades to their left.](figures/fig1_three_way_density.pdf){#fig-density width=100%}

Three things follow.

1. **The two measurements agree.** Flow cytometry over measured $\mathrm{OD}_{600}$ has median
   `r sprintf("%.2f", rat("ratio_FC_over_measuredOD"))`. Two entirely independent
   instruments, one of which needs no dilution assumption, land within about 26% of
   each other at the median.
2. **`N_inoc` agrees with neither**, and is low by a median
   `r sprintf("%.0f", rat("ratio_FC_over_Ninoc"))`× against flow cytometry and
   `r sprintf("%.0f", rat("ratio_measuredOD_over_Ninoc"))`× against $\mathrm{OD}_{600}$.
3. **The vials were not at the stated $\mathrm{OD}_{600}$.** The recorded OD at aliquoting is a
   median `r sprintf("%.2f", rat("ratio_measuredOD_over_statedOD"))`× the paper's
   stated 0.0005, ranging
   `r sprintf("%.2f", a2s$min[a2s$ratio=="ratio_measuredOD_over_statedOD"])`–`r sprintf("%.2f", a2s$max[a2s$ratio=="ratio_measuredOD_over_statedOD"])`×.

## Where does `N_inoc` come from? Not recovered.

```{r a3}
a3p |>
  dplyr::transmute(`Candidate predictor` = candidate, n,
                   `Implied factor (median)` = sci(implied_factor_median, 3),
                   `CV of the factor` = sprintf("%.3f", implied_factor_CV),
                   `Fold spread` = sprintf("%.0f", implied_factor_fold_spread),
                   `log-log r` = ifelse(is.na(pearson_r_log_log), "--",
                                        sprintf("%.3f", pearson_r_log_log))) |>
  kbs(caption = "If N_inoc were built from a candidate by a constant factor, the implied factor would be constant (CV = 0). Sorted by CV; none is close.")
```

The best candidate, the per-replicate `FC_Initial`, still leaves the implied
factor spread over
`r sprintf("%.0f", a3p$implied_factor_fold_spread[1])`-fold (CV
`r sprintf("%.2f", a3p$implied_factor_CV[1])`). `OD_Initial` explains nothing
(log-log $r$ = `r sprintf("%.4f", a3p$pearson_r_log_log[a3p$candidate=="OD_Initial"])`).
`N_inoc` is per-replicate, not per-taxon: all
`r sum(a3s$n_distinct_Ninoc == 5)` taxa have 5 distinct values.

**And it carries a signature no measurement produces.** In
`r sum(abs(a3s$Ninoc_fold_within_taxon - 3) < 1e-5)` of the 15 taxa the largest
`N_inoc` is exactly 3.000000× the smallest, agreeing to 7–8 significant figures:

```{r a3-three}
a3s |> dplyr::filter(abs(Ninoc_fold_within_taxon - 3) < 1e-5) |>
  dplyr::transmute(Taxon, `min N_inoc` = format(Ninoc_min, big.mark = ","),
                   `max N_inoc` = format(Ninoc_max, big.mark = ","),
                   `max / min` = sprintf("%.8f", Ninoc_fold_within_taxon)) |>
  kb(caption = "Taxa whose N_inoc range is exactly a factor of three.")
```

Taken together — no measured quantity reproduces it, and it has an exact
arithmetic structure — **the provenance of `N_inoc` could not be recovered from
the deposit.** That is a finding, not a failure of effort, and it is the first
item in §6.

## `Cell_Counts.csv`, and the parallel-vial reference density

**Confirmed for 15 of 15 taxa.** `data/Cell_Counts.csv` `FC_Initial` is constant
across replicates and equals the rounded arithmetic mean of the per-replicate
values in `data/OD_r_FC_r.csv`. The two files are otherwise identical:

```{r a1}
a1c |> dplyr::transmute(Column = quantity, `Rows` = n_rows,
                        `Identical` = n_identical, `Differing` = n_differing) |>
  kb(caption = "data/Cell_Counts.csv against data/OD_r_FC_r.csv, row by row.")
```

So `Cell_Counts.csv` `FC_Initial` is **a summary of the at-inoculation counts,
not the parallel-vial 45-minute reference density the paper describes**. Searching
the deposit for that parallel-vial measurement finds nothing: no file, column or
timestamp corresponds to it. It exists in the methods and not in the data.

D1 separately established that `data/Cell_Counts.csv` is read by no automated
stage at all — only by the `04` Shiny app, as a fallback list of taxon names.

## The effect of the taxon-mean-versus-per-replicate choice

```{r a4}
a4 |>
  dplyr::transmute(Variant = stringr::str_trunc(variant, 46), n,
                   Intercept = sprintf("%.5f", lmer_intercept),
                   Slope = sprintf("%.4f", lmer_slope),
                   `R2 (mixed)` = sprintf("%.4f", R2_lmer_fitted),
                   `R2 (OLS)` = sprintf("%.4f", R2_ols),
                   `BA bias` = sprintf("%.6f", BA_bias),
                   `% within LoA` = sprintf("%.1f", BA_pct_within)) |>
  kbs(caption = "Growth-rate agreement against the O2 method. Row 2 is what 06_main_figures.R actually computes, and it reproduces D1's growth_rates_combined.csv to exactly 0.",
      align = "lrrrrrrr")
```

The brief anticipated that the published validation used the taxon-mean
initial, putting all replicate-level variation in the numerator. **It does not.**
The counterfactual is worth reporting anyway, because it shows the published
choice is the right one: switching to the taxon mean would drop $R^2$ from
`r sprintf("%.3f", a4$R2_lmer_fitted[3-1])` to
`r sprintf("%.3f", a4$R2_lmer_fitted[3])`, move the slope from
`r sprintf("%.3f", a4$lmer_slope[2])` to `r sprintf("%.3f", a4$lmer_slope[3])`,
and raise the Bland–Altman bias from
`r sprintf("%.6f", a4$BA_bias[2])` to `r sprintf("%.6f", a4$BA_bias[3])` min$^{-1}$.

# Part B — the two back-calculations

## Test the shared assumption first

Both routes stand or fall on exponential growth at the oxygen-derived $r$ over
their interval. That is testable here, because the flow-cytometry fold change is
an independent measurement of the same thing.

```{r b1-sum}
tibble::tibble(Statistic = names(b1s), Value = sprintf("%.4f", as.numeric(b1s))) |>
  kb(caption = "FC_Final / (FC_Initial x exp(r x duration)), all 75 replicates.")
```

**The assumption holds.** The ratio has median
`r sprintf("%.3f", g("ratio median"))`, IQR
`r sprintf("%.3f", g("ratio 25th pct"))`–`r sprintf("%.3f", g("ratio 75th pct"))`,
and `r as.integer(g("n within a factor of 2"))` of
`r as.integer(g("n replicates"))` replicates sit within a factor of two. The
median $r_{FC}/r_{O_2}$ is `r sprintf("%.3f", g("median r_FC / r_O2"))`. This is
also an independent validation of the oxygen-derived growth rate itself.

One replicate is excluded from the backward route, with the reason recorded:

```{r b1-excl}
b1 |> dplyr::filter(!keep_backward) |>
  dplyr::transmute(Taxon, Replicate,
                   `Observed fold` = sprintf("%.2f", fold_observed),
                   `Predicted fold` = sprintf("%.2f", fold_predicted),
                   Ratio = sprintf("%.3f", growth_check_ratio),
                   Reason = "growth saturated or cells died") |>
  kb(caption = "Excluded from the backward route (kept if the ratio is within a factor of 2).")
```

A stricter factor-of-1.5 window would exclude
`r as.integer(g_strict <- b1s[["n excluded (strict, factor 1.5)"]])` replicates
instead of `r as.integer(b1s[["n excluded (primary, factor 2)"]])`; the
conclusions below are unchanged either way.

![(A) The shared assumption: observed against predicted fold change, log-log, with the 1:1 line. (B) The disagreement between the two routes against the fitted growth rate.](figures/fig2_growth_check_and_ratio.pdf){#fig-check width=100%}

## Getting $t_{\text{end}}$ right

The deposit gives two irreconcilable readings, and this matters — they differ by
almost an order of magnitude in $N_0$:

- **End of run.** The paper says endpoint samples were taken "from the
  corresponding SensorDish vials at the end of the run". The oxygen runs end at
  197–212 min. The endpoint sample time itself is not recorded.
- **The recorded interval.** `data/OD_r_FC_r.csv` carries a `Time` column of
  101–139 min that `06_main_figures.R` uses as the initial-to-final interval.

They were tie-broken by a test that uses no endpoint at all: project the
**measured initial** count forward to the fit-window start,
$N_0 = N_{\text{start}}\,e^{r t_0}$, and ask which backward estimate agrees.

```{r b2-tend}
tibble::tibble(
  `Reading of t_end` = c("the recorded OD/FC interval (101-139 min)",
                         "the end of the oxygen run (197-212 min)"),
  `Median agreement with the FC-start projection` = sprintf(
    "%.3f", c(g("CHECK backward / from_FCstart (median, should be 1)"),
              g("CHECK backward_alt / from_FCstart (median, should be 1)")))
) |> kb(caption = "Tie-breaking t_end. A value of 1 means the backward estimate agrees with a projection that uses no endpoint at all.")
```

The recorded interval is therefore used as the primary $t_{\text{end}}$ and end-of-run
is reported as the sensitivity. **Sensitivity:** using end-of-run instead would
give a backward/forward ratio of
`r sprintf("%.1f", g("ratio with t_end = last oxygen reading (median, SENSITIVITY)"))`
rather than `r sprintf("%.0f", g("ratio backward/forward MEDIAN"))` — still far
from 1, so the direction of the finding does not depend on this choice, only its
size.

## Both estimates, and their ratio

```{r b2-sum}
tibble::tibble(Quantity = names(b2s),
               Value = ifelse(abs(as.numeric(b2s)) >= 1e5,
                              sci(as.numeric(b2s), 3),
                              sprintf("%.3f", as.numeric(b2s)))) |>
  kbs(caption = "Both N0 routes, all replicates.")
```

```{r b2-taxon}
b2t |>
  dplyr::transmute(Taxon, n,
                   `N0 forward` = sci(N0_forward_med, 2),
                   `N0 backward` = sci(N0_backward_med, 2),
                   `Ratio min` = sprintf("%.1f", ratio_min),
                   `Ratio median` = sprintf("%.1f", ratio_med),
                   `Ratio max` = sprintf("%.1f", ratio_max),
                   `r (min^-1)` = sprintf("%.4f", r_med)) |>
  kbs(caption = "Backward over forward, by taxon, ordered by the median ratio. Sorting by ratio also sorts by growth rate.",
      align = "lrrrrrrr")
```

Backward over forward has median
**`r sprintf("%.0f", g("ratio backward/forward MEDIAN"))`**, IQR
`r sprintf("%.0f", g("ratio backward/forward 25th pct"))`–`r sprintf("%.0f", g("ratio backward/forward 75th pct"))`,
range `r sprintf("%.1f", g("ratio backward/forward min"))`–`r sprintf("%.0f", g("ratio backward/forward max"))`.
All `r as.integer(g("n with ratio > 10"))` retained replicates exceed 10, and
none is near 1.

**What this measures, rather than bounds.** The forward route's $\exp(r\delta)$
correction has median
`r sprintf("%.2f", g("forward amplification exp(r*delta) median"))`, and Part B's
growth check says growth over $\delta$ really is exponential at that rate. So the
correction is the right *size*. The gap is almost entirely in the starting value:
`N_inoc` is the median `r sprintf("%.0f", rat("ratio_FC_over_Ninoc"))`× below the
measured count found in Part A, and
$`r sprintf("%.0f", rat("ratio_FC_over_Ninoc"))` \times e^{r t_0}/e^{r\delta}
\approx `r sprintf("%.0f", g("ratio backward/forward MEDIAN"))`$. **The forward
correction is not too large or too small; the number it is applied to is wrong.**

## And the disagreement is not a constant

```{r b2-r}
b2r |> dplyr::transmute(Model = model,
                        Intercept = sprintf("%.3f", intercept),
                        Slope = sprintf("%.2f", slope),
                        `SE` = sprintf("%.3f", slope_se),
                        `p` = sci(p_value, 2),
                        `R2` = sprintf("%.3f", r_squared),
                        `Spearman rho` = ifelse(is.na(spearman_rho), "--",
                                                sprintf("%.3f", spearman_rho))) |>
  kb(caption = "log(backward / forward) against the fitted growth rate.")
```

$\log(\text{ratio})$ against $r$ has slope
`r sprintf("%.1f", b2r$slope[1])` min ($p$ = `r sci(b2r$p_value[1], 2)`,
$R^2$ = `r sprintf("%.3f", b2r$r_squared[1])`, Spearman
$\rho$ = `r sprintf("%.3f", b2r$spearman_rho[1])`). Fast-growing taxa disagree by
100–200×, slow-growing ones by about 30×. **Correcting $N_0$ would therefore change
both the level of $R$ and its dependence on $r$** — and it is that dependence
that sets between-taxon respiration activation energies. This is the
method-level finding, not a quirk of this dataset.

## What it does to $R$ and CUE

```{r b3}
b3t |>
  dplyr::transmute(Taxon, n,
                   `R forward` = sprintf("%.0f", R_C_fwd_med),
                   `R backward` = sprintf("%.1f", R_C_bwd_med),
                   `R ratio` = sprintf("%.4f", R_ratio_med),
                   `G` = sprintf("%.1f", G_C_med),
                   `CUE fwd` = sprintf("%.3f", CUE_fwd_med),
                   `CUE bwd` = sprintf("%.3f", CUE_bwd_med)) |>
  kbs(caption = "Per-cell respiration R and growth G in fg C per cell per hour, and CUE, under each route. G does not depend on N0.",
      align = "lrrrrrrr")
```

```{r b3-scale}
b3e |> dplyr::transmute(Effect = effect,
                        `Fold change in R` = sprintf("%.2f", fold_change_in_R),
                        Direction = direction) |>
  kbs(caption = "Every effect on R that D1 and D2 have measured, on one scale.")
```

Set against D1's sensitivity: the **fit-window choice moves $R$ by at most a
factor of 1.23; the $N_0$ route moves it by
`r sprintf("%.0f", b3e$fold_change_in_R[b3e$effect=="N0 route: forward -> backward, effect on R (median)"])`.**
The two are not comparable, and the window sensitivity is not the binding
uncertainty in this analysis.

# Part C — the published temperature result under three routes

The temperature experiment has **no endpoint biomass** — `data/Oxygen_Data_Filtered_CUE.csv`
carries only `Taxon, Temperature, Replicate, Time, Oxygen` — so a true backward
route cannot be computed from it. *Pseudomonas* appears in both datasets, so the
backward route transfers that taxon's measured backward/forward ratio,
`r sprintf("%.2f", PSEUDO_SCALE)`. The third route, $N_0 = N_{\text{inoc}}$ with no
back-calculation at all, is retained as a **lower bound and labelled as such,
not as a candidate**.

```{r c4}
c4 |>
  dplyr::transmute(Route = route,
                   `Growth Topt` = sprintf("%.2f", Topt_growth_rate_C),
                   `Respiration Topt` = sprintf("%.2f", Topt_respiration_C),
                   `CUE Topt` = sprintf("%.2f", Topt_CUE_C),
                   `CUE - growth` = sprintf("%.2f", gap_CUE_minus_growth_C),
                   `shift vs D1 (growth)` = sci(shift_vs_D1_growth, 1),
                   `shift vs D1 (resp)` = sci(shift_vs_D1_resp, 1),
                   `shift vs D1 (CUE)` = sprintf("%+.3f", shift_vs_D1_CUE)) |>
  kbs(caption = "All optima, in degrees C, under the three N0 routes, against D1's published-route baseline.",
      align = "lrrrrrrr")
```

```{r c1}
c1 |>
  dplyr::transmute(Route = route, Response = response, `Best model` = best_model,
                   `Topt` = sprintf("%.2f", Topt_C),
                   `E (eV)` = sprintf("%.3f", E_eV),
                   `E 95% CI` = sprintf("[%.3f, %.3f]", E_lwr_eV, E_upr_eV),
                   `Eh (eV)` = sprintf("%.3f", Eh_eV)) |>
  kbs(caption = "The Sharpe-Schoolfield / Arrhenius parameters written by the pipeline under each route.",
      align = "lllrrlr")
```

**Growth and respiration $T_{opt}$ do not move at all** — the shifts are
`r sci(max(abs(c4$shift_vs_D1_growth)), 1)` and
`r sci(max(abs(c4$shift_vs_D1_resp)), 1)` °C, i.e. numerical noise. This is
structural rather than lucky: the published parameter table fits growth $T_{opt}$
to `growth_r_per_hr`, the raw growth rate, which contains no $N_0$; and
respiration in `10_temperature_cue.R` is $K \cdot O_2^{\text{ref}}$ in carbon
units, which never contained $N_0$ either. **Only CUE can move**, and it moves by
`r sprintf("%.2f", abs(c4$shift_vs_D1_CUE[c4$route=="backward"]))` °C (backward)
and `r sprintf("%.2f", abs(c4$shift_vs_D1_CUE[c4$route=="none"]))` °C (lower
bound).

For completeness, Fig 7 panel A plots growth in *carbon* units, which does
contain $N_0$. Its optimum is at the edge of the measured range under every
route (`r sprintf("%.0f", c3$Topt_C[1])` °C, best model
`r c3$best_model[1]` — a monotonic Arrhenius, i.e. no interior optimum), so it
too is unmoved.

```{r c2}
c2 |>
  dplyr::select(route, Temperature, N0_cells_per_mL, growth_C_fg_per_hr,
                resp_C_fg_per_hr, resp_to_growth_C, CUE) |>
  dplyr::transmute(Route = route, `T (C)` = Temperature,
                   `N0 (cells/mL)` = sci(N0_cells_per_mL, 2),
                   `Growth (fg C/h)` = sci(growth_C_fg_per_hr, 2),
                   `Resp (fg C/h)` = sci(resp_C_fg_per_hr, 2),
                   `Resp:growth` = sprintf("%.2f", resp_to_growth_C),
                   CUE = sprintf("%.4f", CUE)) |>
  kbs(caption = "Per-temperature means, four replicates each, under all three routes.",
      align = "lrrrrrr")
```

![Carbon use efficiency against temperature under the three routes. The level moves by a factor of about 4.5; the optimum (dotted) moves by 0.13 degrees C.](figures/fig3_cue_three_routes.pdf){#fig-cue width=90%}

```{r c4v}
c4v |>
  dplyr::transmute(Route = route,
                   `CUE at 20 C` = sprintf("%.4f", CUE_at_lowest_T),
                   `Peak CUE` = sprintf("%.4f", CUE_at_peak),
                   `T of observed peak` = T_of_observed_CUE_peak,
                   `CUE at 40 C` = sprintf("%.4f", CUE_at_highest_T),
                   `Increases` = ifelse(increases_with_T, "yes", "no"),
                   `Declines at 40 C` = ifelse(declines_at_highest_T, "yes", "no"),
                   `Peaks in 30-35 C` = ifelse(peaks_near_growth_optimum_30_35, "yes", "no"),
                   `Conclusion holds` = ifelse(conclusion_holds, "YES", "no")) |>
  kbs(caption = "Does the paper's stated conclusion survive each route?", align = "lrrrrlllc")
```

What *does* change is the level: peak CUE is
`r sprintf("%.3f", c4v$CUE_at_peak[c4v$route=="forward"])` (forward),
`r sprintf("%.3f", c4v$CUE_at_peak[c4v$route=="backward"])` (backward) and
`r sprintf("%.3f", c4v$CUE_at_peak[c4v$route=="none"])` (lower bound), and the
respiration-to-growth ratio at the optimum goes from
`r sprintf("%.2f", c2$resp_to_growth_C[c2$route=="forward" & c2$Temperature==32])`
to
`r sprintf("%.3f", c2$resp_to_growth_C[c2$route=="backward" & c2$Temperature==32])`.
One honest nuance: under the lower bound the *model-free* per-temperature CUE
peak moves from 32 °C to
`r c4v$T_of_observed_CUE_peak[c4v$route=="none"]` °C even though the fitted
optimum does not. Under the backward route — the actual candidate — both are
unchanged.

## Bottom line {.unnumbered}

```{=latex}
\begin{verdictbox}
```

**No. The published *Pseudomonas* temperature result does not change materially
under a better-supported $N_0$.** Growth $T_{opt}$ stays at
`r sprintf("%.2f", c4$Topt_growth_rate_C[c4$route=="backward"])` °C and
respiration $T_{opt}$ at
`r sprintf("%.2f", c4$Topt_respiration_C[c4$route=="backward"])` °C — identical
to eight significant figures, because neither quantity contains $N_0$. The CUE
optimum moves from `r sprintf("%.2f", c4$Topt_CUE_C[c4$route=="forward"])` °C to
`r sprintf("%.2f", c4$Topt_CUE_C[c4$route=="backward"])` °C, a shift of
`r sprintf("%.2f", abs(c4$shift_vs_D1_CUE[c4$route=="backward"]))` °C, and remains
`r sprintf("%.2f", abs(c4$gap_CUE_minus_growth_C[c4$route=="backward"]))` °C below
the growth optimum and inside the stated 30–35 °C band. CUE still rises with
temperature, peaks at the growth optimum and falls at 40 °C under all three
routes, including the lower bound. **The conclusion as stated in the paper is
robust to the $N_0$ question.**

What is *not* robust is the absolute magnitude. Peak CUE is
`r sprintf("%.2f", c4v$CUE_at_peak[c4v$route=="forward"])` as published and
`r sprintf("%.2f", c4v$CUE_at_peak[c4v$route=="backward"])` under the backward
route, and per-cell respiration across the bacterial set falls by a median factor
of `r sprintf("%.0f", b3e$fold_change_in_R[b3e$effect=="N0 route: forward -> backward, effect on R (median)"])`.
Any statement about how much carbon a cell respires, or about the absolute value
of CUE, rests on `N_inoc` — whose provenance could not be recovered, and which
sits `r sprintf("%.0f", rat("ratio_measuredOD_over_Ninoc"))`–`r sprintf("%.0f", rat("ratio_FC_over_Ninoc"))`×
below both independent measurements in the deposit. That is the part that
warrants a conversation with the authors, and it does not require a correction to
the temperature figure.

```{=latex}
\end{verdictbox}
```

# Questions for the authors {#sec-questions}

Everything here is something the deposited data and the published methods could
not settle.

1. **What is `N_inoculation_cells_per_L`, and how was it produced?** It matches
   no measured quantity in the deposit (best candidate leaves a
   `r sprintf("%.0f", a3p$implied_factor_fold_spread[1])`-fold spread in the
   implied factor), sits
   `r sprintf("%.0f", rat("ratio_measuredOD_over_Ninoc"))`–`r sprintf("%.0f", rat("ratio_FC_over_Ninoc"))`×
   below both independent measurements, and in
   `r sum(abs(a3s$Ninoc_fold_within_taxon - 3) < 1e-5)` of 15 taxa has a
   within-taxon range of exactly 3.000000. This is the single most consequential
   unknown: it scales every absolute respiration number in the paper.
2. **What unit does the CytoFLEX S output represent here** — events per $\mu$L of the
   diluted sample, or already dilution-corrected? The whole flow-cytometry arm of
   Part A rests on this, though the OD cross-check does not.
3. **Where are the parallel-vial 45-minute reference densities?** The methods
   describe them; no file, column or timestamp in the deposit corresponds to
   them. `Cell_Counts.csv` is *not* them — it is the rounded taxon mean of the
   at-inoculation counts (confirmed 15/15).
4. **When were the endpoint $\mathrm{OD}_{600}$/FC samples actually taken?** The text says at
   the end of the run (197–212 min); `OD_r_FC_r.csv` carries an interval of
   101–139 min. The two give $N_0$ estimates a factor of about 10 apart.
5. **Why is the recorded $\mathrm{OD}_{600}$ at aliquoting a median
   `r sprintf("%.1f", rat("ratio_measuredOD_over_statedOD"))`× the stated 0.0005**,
   ranging up to `r sprintf("%.1f", a2s$max[a2s$ratio=="ratio_measuredOD_over_statedOD"])`×?
6. **Is there a taxon-specific $\mathrm{OD}_{600}$-to-cells calibration** for these 15
   isolates? Part A had to use a generic E. coli factor for all of them.
7. **Was the deposited `Oxygen_r` in `OD_r_FC_r.csv` produced by the pipeline in
   this repository?** It differs from both the modular and the original-script
   growth rates (e.g. *Acinetobacter* R1: 0.019508 in the file, 0.019436 modular,
   0.019498 original).

# Which findings are method-level {.unnumbered}

**Method-level** — these apply to the published analysis *and* to the *Candida*
application of the same method, and to any future dataset:

- The forward route makes $\log R$ an explicit decreasing function of the fitted
  $r$. Whenever $N_{\text{inoc}}$ is uncertain, that uncertainty enters $R$
  *correlated with $r$*, not as a constant offset (§3.5: slope
  `r sprintf("%.1f", b2r$slope[1])` min, $R^2$ = `r sprintf("%.3f", b2r$r_squared[1])`).
  Between-taxon orderings of respiration activation energy are therefore not
  robust to it.
- Growth and respiration thermal optima, as this pipeline computes them, are
  *structurally* independent of $N_0$. Only CUE can move. Any $N_0$ debate is a
  debate about CUE levels and about $R$, never about $T_{opt}$ for growth or
  respiration.
- Where an endpoint biomass measurement exists, the back-projection can be
  **measured** rather than bounded — and the measurement here says the
  correction's *form* was right and its *anchor* was not.

**Dataset-specific** — these are about this deposit only:

- The `N_inoc` provenance gap and its exact-factor-of-three signature.
- The missing parallel-vial reference densities.
- The $t_{\text{end}}$ ambiguity.
- The recorded $\mathrm{OD}_{600}$ being about 2.9× the stated value.

# Provenance {.unnumbered}

```{r prov}
tibble::tibble(
  Field = c("Report", "Git commit (environment record)", "Branch",
            "Environment recorded (UTC)", "R", "Platform", "OS",
            "renv lockfile", "Quarto", "PDF engine",
            "Baseline compared against", "Route output trees"),
  Value = c("D2 — N0 reconciliation and dual route",
            VERS$git$commit, VERS$git$branch, VERS$recorded_utc,
            VERS$R$version, VERS$R$platform, VERS$R$os,
            paste0("renv.lock — ", VERS$renv$n_packages, " packages, R ",
                   VERS$renv$lockfile_R),
            VERS$quarto$version,
            paste0(VERS$pdf_engine$name, " — ",
                   stringr::str_trunc(VERS$pdf_engine$version, 44)),
            "runs/D1_baseline (D1, verified byte-identical)",
            "runs/D2_forward, runs/D2_backward, runs/D2_none")
) |> kb(caption = "Provenance. Environment fields come from env/versions.json.")
```

## Where every number came from {.unnumbered}

```{r sources}
tibble::tribble(
  ~Section, ~Source,
  "§2.1 conversion assumptions", "runs/D2_analysis/A2_conversion_assumptions.csv",
  "§2.2 three-way table",        "runs/D2_analysis/A2_three_way_density_by_taxon.csv, A2_three_way_density_per_replicate.csv",
  "§2.2 pairwise ratios",        "runs/D2_analysis/A2_pairwise_ratio_summary.csv",
  "§2.3 N_inoc provenance",      "runs/D2_analysis/A3_Ninoc_provenance_candidates.csv, A3_Ninoc_structure_by_taxon.csv",
  "§2.4 Cell_Counts check",      "runs/D2_analysis/A1_CellCounts_is_taxon_mean.csv, A1_CellCounts_vs_ODFC_columns.csv",
  "§2.5 r_FC recomputation",     "runs/D2_analysis/A4_agreement_published_vs_perrep.csv",
  "§3.1 growth check",           "runs/D2_analysis/B1_growth_check_summary.csv, B1_exponential_growth_check.csv",
  "§3.2 t_end tie-break",        "runs/D2_analysis/B2_N0_ratio_summary.csv",
  "§3.3 N0 both routes",         "runs/D2_analysis/B2_N0_ratio_summary.csv, B2_N0_ratio_by_taxon.csv",
  "§3.4 ratio against r",        "runs/D2_analysis/B2_N0_ratio_vs_r.csv",
  "§3.5 R and CUE",              "runs/D2_analysis/B3_R_and_CUE_by_taxon.csv, B3_effect_sizes_vs_window_choice.csv",
  "§4 optima, three routes",     "runs/D2_analysis/C4_optima_three_routes.csv, C1_thermal_parameters_three_routes.csv",
  "§4 per temperature",          "runs/D2_analysis/C2_per_temperature_three_routes.csv",
  "§4 Fig 7 panel A optimum",    "runs/D2_analysis/C3_growth_carbon_Topt_three_routes.csv",
  "§4 conclusion test",          "runs/D2_analysis/C4_conclusion_holds_three_routes.csv",
  "Pseudomonas transfer scale",  "runs/D2_analysis/C0_pseudomonas_scale.txt",
  "D1 window sensitivity (1.23x)", "runs/D1_baseline/comparisons/D2_per_quantity.csv",
  "Figures 1-3",                 "reports/D2_n0_routes/figures/, from d2_figures.R",
  "Provenance",                  "env/versions.json"
) |> kbs(caption = "Number-to-source map.")
```

Artefacts are produced by `reports/D2_n0_routes/scripts/d2_partA_reconcile.R`,
`d2_partB_routes.R`, `d2_partC_temperature.R` and `d2_figures.R`. The three route
trees come from `10_temperature_cue.R` run with `OXYMODEL_N0_ROUTE` set; the
forward run reproduces `runs/D1_baseline` byte-identically, which is how the
"no default was changed" claim is verified.

# References {.unnumbered}

::: {#refs}
:::
