---
# =============================================================================
# reports/_shared/_template.qmd - skeleton for every D-series findings report
# =============================================================================
# COPY this file to reports/<ID>_<name>/<ID>_<name>.qmd and fill it in.
# Do not render this template itself; it has no content.
#
# The rule (see reports/README.md): a prompt that performs a SUBSTANTIVE
# ANALYSIS writes one of these, rendered to PDF, with the .qmd committed beside
# the PDF. A prompt that produces only reference material writes plain markdown.
# =============================================================================
title: "<ID>: <one-line title>"
subtitle: "<what question this report answers>"
author: "OxygenModel D-series"
date: today
date-format: "YYYY-MM-DD"
format:
  pdf:
    documentclass: scrartcl
    papersize: a4
    geometry:
      - top=25mm
      - bottom=25mm
      - left=22mm
      - right=22mm
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    fig-pos: "H"
    include-in-header: ../_shared/header.tex
bibliography: ../_shared/references.bib
csl: null
execute:
  echo: false
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    dev: "pdf"
---

```{r setup}
# Every report resolves the project root the same way, and reads only
# artefacts that were written by a script - never recomputing an analysis
# inside the report itself.
suppressPackageStartupMessages({
  library(tidyverse); library(knitr); library(jsonlite)
})

BASE <- normalizePath("../..", mustWork = TRUE)
RUN  <- file.path(BASE, "runs", "<run-id>")
rd   <- function(p) suppressWarnings(suppressMessages(
  readr::read_csv(p, show_col_types = FALSE, progress = FALSE)))
VERS <- jsonlite::fromJSON(file.path(BASE, "env", "versions.json"))
```

# Verdict {.unnumbered}

::: {.content-visible when-format="pdf"}
```{=latex}
\begin{verdictbox}
```
:::

*One paragraph. State the answer to the report's question plainly, with the
number that carries it. If the answer is no, say what is missing.*

::: {.content-visible when-format="pdf"}
```{=latex}
\end{verdictbox}
```
:::

# What was done

*What ran, on what inputs, under what environment. Enough that a reader who has
not followed this work can place the report.*

# Findings

*The substance. Every claim carries an actual number. Tables come from the
artefact CSVs, not from numbers typed into the prose.*

# Blockers

*Everything that had to change to make this run, and anything that still cannot
be run. Empty is an acceptable answer, but say so explicitly.*

# Provenance {.unnumbered}

```{r provenance}
tibble::tibble(
  Field = c("Report", "Git commit", "Branch", "Run date (UTC)",
            "R", "Platform", "OS", "renv lockfile", "Quarto", "PDF engine"),
  Value = c("<ID>",
            VERS$git$commit, VERS$git$branch, VERS$recorded_utc,
            VERS$R$version, VERS$R$platform, VERS$R$os,
            paste0("renv.lock (", VERS$renv$n_packages, " packages, R ",
                   VERS$renv$lockfile_R, ")"),
            VERS$quarto$version,
            paste0(VERS$pdf_engine$name, " - ", VERS$pdf_engine$version))
) |> knitr::kable(booktabs = TRUE, linesep = "")
```

## Where every number came from {.unnumbered}

*A table mapping each section of this report to the file its numbers were read
from. No number in a D-series report is allowed to have no source.*

# References {.unnumbered}

::: {#refs}
:::
